Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-01-07 scagnostics.Rmd
Some shuffling in the introduction, removing duplicated bibtex entry.

abstract: |
  Projection pursuit is used to find interesting low-dimensional projections of high-dimensional data by optimizing an index over all possible projections. Most indexes have been developed to detect departure from known distributions, such as normality, or to find separations between known groups. Here, we are interested in finding projections revealing potentially complex bivariate patterns, using new indexes constructed from scagnostics and a maximum information coefficient, with a purpose to detect unusual relationships between model parameters describing physics phenomena. The performance of these indices is examined with respect to ideal behaviour, using simulated data, and then applied to several physics problems. The implementation builds upon the projection pursuit tools available in the R package, tourr, with indices constructed from code in the R package, scagnostics.

@@

A large number of projection pursuit indices have been developed, to detect departure from multivariate normal, which includes clusters or outliers or separations between known groups (e.g. @f87, @CBC92, @lckl2005, @AHC02, @CEM:CEM2568, @JS87, @5508437).  Less work has been done on indexes to find nonlinear dependence between variables.

XXX Need to add something here about the guided tour

@@

A related topic is variable selection. With high-dimensional data, even plotting all pairs can be daunting, so "scagnostics" [@scag] [@WW08] have been developed to find the most interesting pairs of variables. There are 8 scagnostics, vividly named: "Outlying", "Skewed", "Sparse", "Clumpy", "Striated", "Convex", "Skinny" and "Stringy". Our question is whether these can be extended into projection pursuit indexes that can be used to find two-dimensional projections of high-dimensional data with unusual features.

@@

<!--
Suggestion German:
We are motivated by physics applications, specifically to assist the interpretation of models fit to experimental results. In these problems one has a physical model with a set of m free parameters which cannot be measured directly, but that are determined instead by fitting a set of p experimental observations. The model predicts all the observations in terms of the m parameters, often analytically, but occasionally only numerically. Moreover, a single prediction can be a complicated function of all the free parameters requiring intensive numerical resources.

These problems often deal with a large number of observations and parameters, with p in the hundreds or even thousands and m of order 10. 
The results of a fit have been generally interpreted in terms of combinations of parameters selected by intuition or prior knowledge. Recently, these selections have been also assisted by machine learning techniques including BDT and neural networks.

Our visualisation tools complement these procedures providing a high dimensional picture which assists with the interpretation of the results,  serves as a check for other methods and possibly revealing new associations.
-->

[@Grimm2016] explored the behavior of scagnostics for selecting variables, and proposed two more, based on smoothing splines and distance correlation, that have some nicer properties. In addition the availability of another two indices based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], and computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308], round out the collection of metrics. \footnote{It is interesting to note that entropy, closely related to the notion of mutual information, has previously been considered as a projection pursuit index when searching for interesting one-dimensional projections, see discussion in \cite{huber85,jones87}.}
These indices are all available in R packages [@rref]: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr].

@@

\textcolor{red}{\cite{jones87} discusses centering and sphering of the data points. Centering is needed if index is not translational invariant (should not be the case for any index considered here). Arguments for sphering are that scale effects should be extracted, we don't want to reproduce information already obtained by PCA. To my mind rescaling to an overall interval is more appropriate here. Rotations between parameters should not affect the outcome, but will make interpretation more difficult. Standardizing the variance will remove information about the posterior variance. Overall we are interested in strong association only, if these structures were affected by the standardization procedure, it would most likely be unwelcome effects?}

@@

A useful projection pursuit index needs to have several properties. This has been discussed in several seminal papers (\cite{diaconis84}, \cite{huber85}, \cite{jones87}, \cite{posse95b}, \cite{hall89}). The PPI should be  minimized by the normal distribution, because this is not interesting from a data exploration perspective. If all projections are normally distributed, good modeling tools already exist. A PPI should be approximately affine invariant, regardless how the projection is rotated the index value should be the same, and the scale of each variable shouldn't affect the index value. Interestingly the original index proposed by \cite{FT74} was not rotationally invariant. A consistent index means that small perturbations to the sample do not dramatically change the index value. This is particularly important to making optimization feasible, small angles between projections correspond to small perturbations of the sample, and thus should be small changes to index value. \cite{posse95b} suggests that indexes should be resistant to features in a tail of the distribution, but this is debatable because one departure from normality that is interesting to detect are anomalous observations. Some PPI are designed precisely for these reasons. Lastly, because we need to compute the PPI over many projections, it needs to be fast to compute. These form the basis of the criteria upon which the scagnostic indexes, and the several alternative indexes are examined, as explained below.

>>>>

abstract: |
  Projection pursuit is used to find interesting low-dimensional projections of high-dimensional data by optimizing an index over all possible projections. Most indexes have been developed to detect departure from known distributions, such as normality, or to find separations between known groups. Here, we are interested in finding projections revealing potentially complex bivariate patterns, using new indexes constructed from scagnostics and a maximum information coefficient, with a purpose to detect unusual relationships between model parameters describing physics phenomena. The performance of these indices is examined with respect to ideal behaviour, using simulated data, and then applied to several physics problems. The implementation builds upon the projection pursuit tools available in the R package, tourr, with indexes constructed from code in the R packages, scagnostics, minerva and mbgraphics.

@@

A large number of projection pursuit indexes have been developed, to detect departure from multivariate normal, which includes clusters or outliers or separations between known groups (e.g. @f87, @CBC92, @lckl2005, @AHC02, @CEM:CEM2568, @JS87, @5508437).  Less work has been done on indexes to find nonlinear dependence between variables. \footnote{\textcolor{red}{The second component in implementing a projection pursuit method is the optimisation algorithm. Several algorithms have been suggested, some tailored to specific applications, see e.g. XXX references?. In this work our focus is on the index functions, and we will only use optimisation algorithms available via the guided tour implementation of tourr.}}

@@

@@

We draw on developments from the related topic of variable selection. With high-dimensional data, even plotting all pairs can be daunting, so "scagnostics" [@scag] [@WW08] have been developed to find the most interesting pairs of variables. There are 8 scagnostics, vividly named: "Outlying", "Skewed", "Sparse", "Clumpy", "Striated", "Convex", "Skinny" and "Stringy". Our question is whether these can be extended into projection pursuit indexes that can be used to find two-dimensional projections of high-dimensional data with unusual features.

[@Grimm2016] explored the behavior of scagnostics for selecting variables, and proposed two more, based on smoothing splines and distance correlation, that have some nicer properties. In addition the availability of another two indices based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], and computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308], round out the collection of metrics. \footnote{It is interesting to note that entropy, closely related to the notion of mutual information, has previously been considered as a projection pursuit index when searching for interesting one-dimensional projections, see discussion in \cite{huber85,JS87}.}
These indices are all available in R packages [@rref]: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr].

@@

\textcolor{red}{\cite{JS87} discusses centering and sphering of the data points. Centering is needed if index is not translational invariant (should not be the case for any index considered here). Arguments for sphering are that scale effects should be extracted, we don't want to reproduce information already obtained by PCA. To my mind rescaling to an overall interval is more appropriate here. Rotations between parameters should not affect the outcome, but will make interpretation more difficult. Standardizing the variance will remove information about the posterior variance. Overall we are interested in strong association only, if these structures were affected by the standardization procedure, it would most likely be unwelcome effects?}

@@

A useful projection pursuit index needs to have several properties. This has been discussed in several seminal papers (\cite{diaconis84}, \cite{huber85}, \cite{JS87}, \cite{posse95b}, \cite{hall89}). The PPI should be  minimized by the normal distribution, because this is not interesting from a data exploration perspective. If all projections are normally distributed, good modeling tools already exist. A PPI should be approximately affine invariant, regardless how the projection is rotated the index value should be the same, and the scale of each variable shouldn't affect the index value. Interestingly the original index proposed by \cite{FT74} was not rotationally invariant. A consistent index means that small perturbations to the sample do not dramatically change the index value. This is particularly important to making optimization feasible, small angles between projections correspond to small perturbations of the sample, and thus should be small changes to index value. \cite{posse95b} suggests that indexes should be resistant to features in a tail of the distribution, but this is debatable because one departure from normality that is interesting to detect are anomalous observations. Some PPI are designed precisely for these reasons. Lastly, because we need to compute the PPI over many projections, it needs to be fast to compute. These form the basis of the criteria upon which the scagnostic indexes, and the several alternative indexes are examined, as explained below.
